### 2022-05-25 src/ms.tex
chapter 2 progress

\section{Gravitational microlensing}
\emph{``Do not bodies act upon light at a distance and by their action bend its rays,
    and is not this action strongest at the least distance?''} asked \citet{Newton1704}.
Much later in 1912, even before he published his theory of General Relativity, Einstein
predicted the that a distant light source could be magnified by a foreground mass
relative to an observer on Earth if the two objects were very precisely aligned
\citep{Renn1997}.
At the time Einstein concluded that this \emph{gravitational lensing} effect was
unobservable and did not even bother
publishing the result until 24 years later a young amateur scientist R.W. Mandl convinced
him to write a short paper on the subject \citep{1936Sci....84..506E}.
In that paper he
stated that \emph{``Of course, there is no hope of observing this phenomenon directly.
    First, we shall scarcely ever approach closely enough to such a central line.
    Second, [the angles] will defy the resolving power of our instruments''}.
Still, the consensus at the time was that this was unobservable.

The first definitive observation came in 1979 by \citet{1980ApJ...241..507Y}
who observed a double image of the quasar Q0957+561 and concluded that the two
images correspond to the same object whose light was distorted by a foreground
gravitational lens in the form of a massive galaxy. That the same effect could
be used to detect the presence of planets orbiting around their host stars was
first theorized by \citet{1964PhRv..133..835L} who wrote that \emph{``the
    primary effect of planetary deflectors bound to stars other than the Sun will
    be to slightly perturb the lens action of these stars''} although he was also
sceptical about the possibility of detection, saying that \emph{``associated
    pulses would be so weak and infrequent and of such fleeting duration – perhaps
    a few hours – as to defy detection''}. Gravitational lensing as a method for
discovering exoplanets really took off with the work of Paczyński
\citep{1986ApJ...304....1P,1986ApJ...301..503P,1991ApJ...374L..37M} who also
coined the term \emph{microlensing}, which refers to gravitational lensing in a
regime where the images of the background object cannot be resolved but one can
nevertheless measure its magnification as a function of time.

@@

\section{Gravitational microlensing}
\label{sec:microlensing}

\subsection{Deflection of light by gravity}

\subsection{The different scales of lensing}
%% Read the "scales of lensing" paper by that Italian guy
\subsection{Microlensing}
\subsection{The point mass lens and the lens equation}
\subsection{Parallax}
\subsection{Finite source effects}
\subsection{The binary lens}
\subsection{Multiple lenses}
\subsection{Solving the lens equation}

>>>>

\section{Gravitational microlensing}
\emph{``Do not bodies act upon light at a distance and by their action bend its rays,
    and is not this action strongest at the least distance?''} asked \citet{Newton1704}.
Much later in 1911, even before he published his theory of General Relativity (GR), Einstein
considered the deflection of light from a distant star passing close
to a massive object in the foreground and derived an expression for the deflection angle of the light ray
which was off by a factor of 2 relative to the correct value.
He corrected the error in 1915 using GR.
The same year Albert Einstein published his General theory of Relativity (1916),
English physicist Sir Oliver Lodge suggested the light bending phenomenon could produce
a \emph{gravitational lens}.
He warned that it wouldn't be a lens in the usual sense,  because it does not have a
focal length.

The first experimental confirmation of the deflection of light by a mass came
in 1919 during the Solar eclipse when the astronomer Arthur Eddington famously
measured the deflection of light from distant stars passing close to the limb
of the Sun \footnote{He concluded that the observed deflection was in agreement
    with GR, although it is doubtful that the data was actually good enough to
    distinguish between the GR prediction for the deflection of light and the
    Newtonian prediction which is derived using the equivalence principle and is
    equal to one half the GR value.}. Much later in 1936 a young amateur scientist
R.W. Mandl convinced Einstein to write a short a paper on the lensing effect by
a massive star acting as the lens instead of the Sun. In the paper
\citep{1936Sci....84..506E} Einstein derived an expression for the
magnification of the distant star when it is closely aligned to the foreground
lens star relative to an observer and he predicted that a luminous ring would
form if the two stars were perfectly aligned. He considered this effect curious
but useless, stating that \emph{``Of course, there is no hope of observing this
    phenomenon directly. First, we shall scarcely ever approach closely enough to
    such a central line. Second, [the angles] will defy the resolving power of our
    instruments''}.

In this matter, Einstein was wrong. The famous astronomer Fritz Zwicky noticed
the phenomenon is likely to be observable at galactic scales if the lens was a
massive galaxy \citep{1937PhRv...51..290Z,1937PhRv...51..679Z} and that one
could use the measurement of the deflection angle to weigh the lens. Zwicky was
right and the first definitive observation came in 1979 by
\citet{1980ApJ...241..507Y} who observed a double image of the quasar Q0957+561
and concluded that the two images correspond to the same object whose light was
distorted by a massive galaxy. Many other observations followed, including
visible Einstein rings. Far from being just a curious phenomenon, galactic
scale lensing is now one of the key methods of observational cosmology, used
for inferring the parameters of the standard model and the distribution of dark
matter in the universe.

@@

\section{Gravitational microlensing}
\label{sec:microlensing}
Gravitational lensing is generally divided into multiple classes depending
on weather the effect is discernible at the level of individual objects or a
a statistical sample of object. The main division is between \emph{weak lensing}
and \emph{strong lensing}.
The former refers to lensing by galaxies and clusters of galaxies on cosmological
scales where the deflection  is impossible to detect in a single background source
but one can tease out the effect in a statistical sense.
Strong lensing refers to lensing of individual objects and it is further divided into
\emph{macrolensing} -- lensing of galaxies where the multiple images of the source
are resolved, and \emph{microlensing}  -- the images are generally not resolved,
and the observable is the total magnification of the source as a function of time (a light curve).
In case of microlensing the light source is either a quasar or a star (sometimes a binary star)
and the lenses are stars, brown dwarfs, planets, and compact objects such as black holes, neutron
stars and white dwarfs. In this thesis we focus specifically on microlensing
with stars as the light source instead of quasars\footnote{Quasar microlensing
    is a somewhat separate community from the rest of the microlensing community.}.

\subsection{Deflection of light by gravity}
Gravitational lensing is a phenomenon fully described by Einstein's General
Theory of Relativity (GR) in which light changes its direction of propagation
when passing close to a massive body. GR predicts the following:
\begin{itemize}
    \item the presences of mass changes the spacetime geometry from the flat (Minkowski)
          metric $\eta_{\mu\nu}$ to a curved metric, specified by the tensor $g_{\mu\nu}$
    \item massless particles such as photons follow the null geodesics -- minimum
          distance paths in curved spacetime which are obtained by solving the
          \emph{geodesic equation}
\end{itemize}
If we restrict ourselves to the regime where the metric is time independent
and the particles are allowed to travel at any velocity less than $c$,
also known as the \emph{weak field} or \emph{linearized approximation }
of GR, the metric tensor
can be written as
\begin{equation}
    g_{\mu\nu}=\eta_{\mu\nu}+h_{\mu\nu}\quad,
\end{equation}
where $h_{\mu\nu}=
    \textrm{diag}(-2\Phi,-2\Phi,-2\Phi,-2\Phi)$\footnote{Assuming the
    the metric sign convention $(-,+,+,+)$} \citep{carroll_2019}.
The (static) Newtonian gravitational potential $\Phi$ obeys the Poisson equation
\begin{equation}
    \nabla^2\Phi=4\pi G\rho\quad,
\end{equation}
where $\rho$ is the mass density of the distribution of mass located between the observer
and the light source.
The geometry of the lensing system consisting of a distant source of light, the
observer, and a distribution of mass with density $\rho$ located between the
source and the observer is shown in Figure~\ref{fig:deflection_angle}.
\begin{figure}
    \centering
    \includegraphics[width=0.6\linewidth]{static/microlensing/deflection_angle.pdf}
    \caption{Geometry of gravitational lensing.
        A geodesic curve $x^\mu(\lambda)$ is deflected
        by an angle $\hat\alpha$ from its initial trajectory
        due to the presence of a massive body with mass density $\rho$. Figure
        adapted from \citet{carroll_2019}.}
    \label{fig:deflection_angle}
\end{figure}
A photon is deflected as it travels from a source to an observer by a
\emph{deflection angle} $\hat\alpha$, a vector in the
plane  perpendicular to the wave vector $\vec{k}$ pointing in the direction of
photon propagation.
The deflection angle is given by \citep{carroll_2019}
\begin{equation}
    \hat\alpha=2\int\nabla_\perp\Phi \ud s\quad,
    \label{eq:deflection_angle_general}
\end{equation}
where $\nabla_\perp\Phi$ is the gradient of the potential in the direction transverse
to the path of the photon and $s$ is the spatial distance traveled.
For a point mass $M$, the potential is given by the potential is
\begin{equation}
    \Phi=- \frac{GM}{(b^2+x^2)^{1/2}}\quad,
\end{equation}
where $x$ parametrizes the straight line distance between the observer and the lens and
$b$ is the impact parameter of the light ray.
Integrating from $-\infty$ to $\infty$ (i.e. assuming that both
the source and the observer are located far away from the deflecting mass), we obtain
the deflection angle:
\begin{equation}
    \hat\alpha= \frac{4GM}{c^2b}= \frac{2R_s}{b}\quad,
\end{equation}
where $R_s=2GM/c^2$ is the Schwarzschild radius.
$\hat\alpha$ is directly proportional to the mass of the lens and
it does not depend on the wavelength of the light. It also increases with the proximity of
the light ray to the lensing mass which is the opposite behavior to that of a
classical convex lens for which the deflection angle vanishes at the center of the lens.
The gravitational lens does not have a focal length and so it is not a lens in the usual
sense.
The resulting deflection angle is very small, for example, for the Sun we have
$GM/c^2=1.48\times 10^5\si{\centi\meter}$ (2.95 km), $R=6.96\times
    10^{10}\si{\centi\meter}$ and $\hat\alpha=1.75$ arc seconds. This is the angle
that was claimed to have been observed by Eddington when during the 1919 total
solar eclipse.

Equation~\ref{eq:deflection_angle_general} is only valid when the linearized
approximation of GR is sufficient. In this approximation, the deflection angles
are additive and the total deflection caused by a group massive bodies is just
the sum of the individual deflection angles. The linearized approximation
breaks down when the impact parameter $b$ approaches the Schwarzschild radius
of the lensing mass and the linearized metric is no longer sufficient. Except
for modeling lensing in the close vicinity of a black, the linearized
approximation is sufficient.

\subsection{The magnification of a point source by a point lens}
Consider a system consisting of an observer $O$, a point mass $M$, and a point
light source $S$. The geometry of this system is shown in
Figure~\ref{fig:lens_geometry}. The lens is located in the \emph{lens plane}, a
plane perpendicular to the observer--lens axis, at a distance $D_L$ away from
the source. The light source is located in the \emph{source plane}
perpendicular to the observer--source axis, at a distance $D_S$ away from the
observer, and the observer is in the \emph{observer plane}. The assumption that
the location of the source, observer, and the lens can be parametrized using
angular coordinates (or Cartesian coordinates if the distances to the lens and
the source are known) in their respective planes is justifiable because the
deflection angles involved are very small and the distance between the observer
and any point of interest in the lens plane is approximately constant.
\begin{figure}[!ht]
    \centering
    \includegraphics[width=0.4\linewidth]{static/microlensing/lens_geometry.pdf}
    \caption{The geometry of a system consisting of a point mass lens $M$ located at distance
        $D_L$ at angular separation $\beta$ from the observer, and a single light source $S$ located at distance $D_S$, emitting a light ray which gets
        deflected by an angle $\hat\alpha$ from its
        original trajectory. As a result the observer sees the source at angular separation
        $\theta$. Figure adapted from \citet{1992grle.book.....S}.}

    \label{fig:lens_geometry}
\end{figure}
Let $\theta$ be the apparent angular position of the source in the sky with respect to the
observer--lens axis and $\beta$ be the actual angular position of the source.
Assuming a metric for spacetime in between the source and the observer that is
approximately Euclidian\footnote{This is a valid assumption for galactic sources but not for
    extragalactric source such as quasars and galaxies which require a cosmological
    model for the metric.} the distance from the lens to the
source is $D_S-D_L$.

From Figure~\ref{fig:lens_geometry} it follows that the relationship between
the observed and the actual location of the source is given by
\begin{equation}
    \beta=\theta-2 R_s \frac{D_S-D_L}{D_LD_S}
    \frac{1}{\theta}\quad,
    \label{eq:lens_equation}
\end{equation}
which is known as the \emph{lens equation} or sometimes also the \emph{ray-tracing}
equation. If the source and the lens are perfectly aligned with respect to the observer,
then $\beta=0$ and we have
\begin{equation}
    \theta \equiv \theta_E= \sqrt{ \frac{4GM}{c^2} \left( \frac{1}{D_L} - \frac{1}{D_S} \right)}
\end{equation}
The distance term $D_L^{-1}-D_\textrm{s}^{-1}$ is proportional to the
lens-source parallax.

The image of the source forms a ring (the Einstein ring) around the lens star
with angular radius $\theta_E$ -- the \emph{Einstein radius}. $\theta_E$
depends on the mass of the lens and the distances to the lens and the source.
The angular Einstein radius sets the characteristic scale of a microlensing
event, it is a natural unit for most microlensing parameters. For an M-dwarf
star lens in the galactic disc ($D_L\sim 3\,\textrm{kpc}$) and a source star in
the galactic bulge ($D_S\sim 8\,\textrm{kpc}$) we have $\theta_E\sim
    1\,\textrm{mas}$ so the typical scale of microlensing is on the order of
miliarcseconds. The images of the source star are only rarely observable using
the most advanced optical interferometers such as the GRAVITY instrument on
ESO's Very Large Telescope whose lower resolution limit is $2\,\textrm{mas}$
\citep{arXiv:1705.02345}. \cite{2019ApJ...871...70D} were the first to resolve
the images of a microlensing event, finding the value of
$\theta_E=1.87\,\textrm{mas}$ using the GRAVITY instrument.

We can rewrite Equation~\ref{eq:lens_equation} by defining new dimensionless
angular coordinates scaled by the angular Einstein radius:
\begin{equation}
    x= \frac{\theta}{D_L\theta_E}, \quad y=\frac{\beta}{D_S\theta_E}
\end{equation}
The lens equation then takes the very simple form
\begin{equation}
    y= x- \frac{ 1}{x}
    \label{eq:lens_equation_dimensionless}
\end{equation}
Equation~\ref{eq:lens_equation_dimensionless} is a nonlinear mapping between the
source plane $y$ and the image plane $x$.
If we solve for $x$, we obtain two solutions for the position of the images, given by

\begin{equation}
    x_\pm= \frac{1}{2} \left( y\pm\sqrt{y^2 +4}\right)
    \label{eq:images_location}
\end{equation}
For non zero $y$ the so called \emph{minor image} is always located
within the Einstein radius ($\lvert x_-\rvert<1$) while the \emph{major image}
is located outside of it ($\lvert x_+\rvert>1$).

Since gravitational lensing does not involve any additional emission or
absorption along a deflected ray of light and since the wavelength of the light
ray does not change when it encounters the lens, the \emph{surface brightness}
(flux density per unit angular area) $I$ of the source image is identical to
the surface brightness of the unlensed source. The flux of an infinitesimal
source is a product of the surface brightness and the solid angle $\Delta
    \omega$ subtended by the source in the sky which changes throughout the
microlensing event because the source and the lens are in motion relative to an
observer. The ratio of the original and the lensed source flux, the
\emph{magnification} $A$, is then given by the ratio of the two solid angles
\begin{equation}
    A= \frac{\Delta \omega}{(\Delta\omega)_0}\quad,
    \label{eq:magnification}
\end{equation}
where $0$ denotes the unlensed solid angle.
Assuming an infinitesimal source size at angular location $\mathbf y =(y_1,y_2)$ subtending an angle
$\Delta (\omega)_0$, and an image of the source at location $\mathbf x=(x_1,x_2)$ subtending an angle
$\Delta \omega$; the ratio between the solid angle of the source and the image
is given by the determinant of the Jacobian matrix of the
lens mapping $\mathbf{x}\rightarrow\mathbf{y}$, evaluated at the location of the images
\begin{equation}
    \frac{(\Delta\omega)_0}{\Delta\omega} =\left\lvert\textrm{det}
    \frac{\partial \mathbf y}{\partial \mathbf x} \right\rvert
\end{equation}
The magnification $A$ is then given by the inverse Jacobian determinant of the lens mapping:
\begin{equation}
    A= \left\lvert\det
    \frac{\partial \mathbf y}{\partial \mathbf x} \right\rvert^{-1} \label{eq:magnification_general}
\end{equation}
The images for which the determinant of the Jacobian of the lens mapping is
positive are said to have positive \emph{parity} and vice versa.

Looking at Equation~\ref{eq:magnification_general}, we see that the
magnification diverges when the Jacobian determinant of the lens mapping
vanishes. Curves in the lens plane for which the determinant of the lens
mapping vanishes, that is,
\begin{equation}
    \left\lvert\det
    \frac{\partial \mathbf y}{\partial \mathbf x} \right\rvert=0
\end{equation}
are called \emph{critical curves}.
The critical curves in the lens plane can be mapped to the source plane using
the lens equation $\mathbf{x}\rightarrow \mathbf{y}$ and these curves are
are called \emph{caustic curves}.
Although the magnification factor in Equation~\ref{eq:magnification_general} formally
diverges at the points of critical curves or caustics,
the divergence is not physical because light sources aren't truly point like.
If the finite angular size of the source is taken into
account, the magnification is an integral of Equation~\ref{eq:magnification_general} over the
extent of the source, weighted by the source brightness and that integral is always finite.
Both the critical and the caustic curves are closed and one can prove that \textbf{the
    number of images changes by two if and only if the source crosses a caustic curve}
\citep[][Chapter 6]{1992grle.book.....S}.

We can derive a simpler expression for Equation~\ref{eq:magnification} by
switching to polar coordinates $(v,\phi)$ in the lens plane. The lens equation
for the two vector components of the apparent source position is given by
\begin{align}
    y_1= & x_1- \frac{x_1}{x_1^2+x_2^2} \\
    y_2= & x_2- \frac{x_2}{x_1^2+x_2^2}
\end{align}
Introducing polar coordinates $x_1=v\cos\phi,\;x_2=v\sin\phi$, we have
\begin{equation}
    \left\lvert\textrm{det}
    \frac{\partial \mathbf y}{\partial \mathbf x} \right\rvert =1- \frac{1}{v^4}\quad,
\end{equation}
By substituting in the locations of the source images Equation~\ref{eq:images_location} into the
above expression, we obtain the magnification of the source for both images
\begin{equation}
    A_\pm=\left(1- \frac{1}{x^4_\pm} \right)^{-1}
\end{equation}
The total magnification is then the sum of the two magnifications, and is given by
\citep{1936Sci....84..506E}
\begin{equation}
    A(u)=\lvert A_-\rvert+\lvert A_+\rvert= \frac{u^2+2}{u\sqrt{u^2+4}}\quad,
    \label{eq:magnification_point_lens}
\end{equation}
where $u = \sqrt{y_1^2+y_2^2}$ is the the magnitude of the position
vector of the source star.
The critical curve of the point lens is the Einstein
ring corresponding to $v=\sqrt{x_1^2+x_2^2}=1$ and the caustic curve is mapped to a
a single point in the source plane at $u=0$.
For source separations much smaller than the angular Einstein radius ($u\ll 1$), the
magnification is approximately $A(u)\simeq 1/u$, in the opposite case ($u\gg 1$)
we have $A(u)\simeq 1+ 2/u^4$, that is, the magnification falls off rapidly the further
away the source is from the lens.

%The angular Einstein radius determines a characteristic timescale
%$t_E=\theta_E/|\boldsymbol{\mu}_{LS}|$ on which the magnification of of the
%source changes, where $\boldsymbol{\mu}_{LS}$ is the relative proper motion
%vector between the source and the lens
%$\boldsymbol{\mu}_{LS}\equiv\boldsymbol{\mu}_L- \boldsymbol{\mu}_S$. The
%predicted flux of a star undergoing a microlensing event is then given by the
%following expression
%\begin{equation}
%    f=F_SA(t)+F_B
%    \label{eq:flux_paczynski}
%\end{equation}
%where $F_S$ is the intrinsic flux of the star and
%$F_B$ is any background flux due to other stars in the field which
%might be blended with the source star, usually called the \emph{blending flux}.
%$A(t)$ is a non-linear time-dependant magnification mapping which depends
%on the properties of the lens and the geometry of the lensing event.
%In the simplest case, that of a point source being magnified by a point lens,
%the magnification is given by
%\begin{equation}
%    A(t)= \frac{u(t)^2+2}{u(t)\sqrt{u(t)^2+4}}
%\end{equation}
%where $u(t)$ is the separation between the lens and the source expressed in
%units of the angular Einstein radius.

To be able to evaluate Equation~\ref{eq:magnification_point_lens} in practice
we have to parametrize the position of the source on the sky $u$ as a function
of time. To derive an expression for $u(t)$ we assume the the motion of the
observer, lens and the source is rectilinear (acceleration can be neglected).
We take $\boldsymbol{\theta}_S$ to be the angular position of the source star
on the plane of the sky and $\boldsymbol{\mu}_L$ to be its proper motion
vector. Likewise for the lens. We thus have
\begin{align}
    \boldsymbol{\theta}_S(t) & =\boldsymbol{\theta}_{S, 0}+
    \left(t-t_{0}\right) \boldsymbol{\mu}_S
    \label{eq:source_position}                              \\
    \boldsymbol{\theta}_L(t) & =\boldsymbol{\theta}_{L, 0}
    +\left(t-t_{0}\right) \boldsymbol{\mu}_L\quad,
    \label{eq:lens_position}
\end{align}
where $t_0$ is some reference time.
The relative position vector of the lens with respect to the source is then
\begin{equation}
    \boldsymbol{u}(t) \equiv \frac{\boldsymbol{\theta}_L(t)
        -\boldsymbol{\theta}_S(t)}{\theta_E}=
    \frac{\boldsymbol{\theta}_{LS, 0}}{\theta_E}
    +\frac{t-t_{0}}{\theta_E} \boldsymbol{\mu}_{LS}
    \label{eq:relative_trajectory_no_parallax}
\end{equation}
where
$\boldsymbol{\theta}_{L S, 0}\equiv\boldsymbol{\theta}_{L,0}-\boldsymbol{\theta}_{S, 0}$
is the relative position at $t_0$ and
$\boldsymbol{\mu}_{LS}\equiv \boldsymbol{\mu}_{L}- \boldsymbol{\mu}_{S}$ is the
relative proper motion.
Since $\boldsymbol{\mu}_{LS}$ and $\boldsymbol{\theta}_{L S, 0}$ are
perpendicular to each other, it follows that the magnitude of the relative
separation is
\begin{equation}
    u(t)=\sqrt{u_0^2+ \left(\frac{t-t_0}{t_E}\right)^2}
\end{equation}
where we have defined $u_0\equiv |\boldsymbol{\theta}_{L S, 0}|/\theta_E$
and $t_E\equiv \theta_E/|\boldsymbol{\mu}_{LS}|$.
The model for the light curve is then characterized by
the parameters $\{F_S,F_B, t_0, u_0, t_E\}$ and is known in the
literature as the Paczy\'nski curve.

The magnification as a function of time for different impact parameters $u_0$
is shown in Figure~\ref{fig:single_lens_mag}. Notice the very steep fall off in
magnification as as we move away from the Einstein ring ($A(u)\propto
    1+2u^{-4}$). The curves shown in Figure~\ref{fig:paczynski_curve} are often
called \emph{Paczyński curves} \citep{1986ApJ...304....1P,1986ApJ...301..503P},
in honour of the late Polish astronomer Bohdan Paczyński, one of the key
figures of gravitational microlensing.

\begin{figure}[t]
    \begin{centering}
        \includegraphics[width=0.7\linewidth]{figures/paczynski_curve.pdf}
        \caption{
            Magnification of a point source by a point lens as a function of time. The
            various magnification curves correspond to different impact parameters $u_0$. The inset figure
            shows the source trajectories in the lens plane, the dashed circle corresponds
            to the Einstein radius $v=1$.}
        \label{fig:paczynski_curve}
    \end{centering}
\end{figure}

\subsection{Magnification of an extended source}
For a relatively small subset of microlensing events the source star passes
very close to the caustic and the variation of magnification over the extent of
the source star disc is non-negligible. For those events the point source
approximation breaks down and we have to take into account the finite angular
radius $\theta_\star$. This will generally happen when $u_0 \lesssim
    \rho_\star/2$ \citep{1997ApJ...477..580G} where
$\rho_\star=\theta_\star/\theta_E$. The effect of finite source effects on the
magnification curve thus matters only near the peak of an event and it results
in a rounder top of the curves shown in Figure~\ref{fig:paczynski_curve}.
Unfortunately there are no analytic solutions for the magnification of an
extended source lensed by a single point lens with a general surface brightness
profile. \citet{1994ApJ...421L..71G} derived an approximate solution valid for
uniform brightness sources with $\rho_\star\lesssim 0.1$ and
\citet{1994ApJ...430..505W} derived a general solution (also for a uniform
brightness source) involving solutions to elliptic integrals of the first,
second and third kind. A general solution involving Fast Fourier Transforms
which is applicable to any source profile was recently proposed by
\citet{2022arXiv220306637S}.

\begin{figure}[t]
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/single_lens_images.pdf}
        \caption{Images of an extended limb-darkened source star lensed by a single
            point lens for varying position of the source star. The top panels show the magnification map with a logarithmic scale
            colormap  consisting of a single point caustic. The semi-transparent circle
            is the source star disc with radius $\rho_\star=0.15$. The bottom row shows
            the two images merging into the Einstein ring as the source moves over the caustic.}
        \label{fig:single_lens_images}
    \end{centering}
\end{figure}

In Figure~\ref{fig:single_lens_images} we show the magnification map and the
images of an extended limb-darkened source with $\rho_\star=0.15$ as it
approaches the point caustic for a single lens. The top panels show the
magnification map in the source plane using a logarithmic scale colormap. The
source disc is shown (semi-transparent grey circle) at different positions
relative to the caustic. The bottom panels show the resulting images in the
image plane, this is what we would see on the sky if we could resolve the
microlensing event. As the source moves closer to the caustic the area (and
hence the magnification) of the two images increases and eventually they merge
forming an Einstein ring.

To produce these plots I evaluated the lens mapping
(Eq.~\ref{eq:lens_equation_dimensionless}) on a regular grid in the image plane
and then plotted the surface brightness of the corresponding point on the
source disc in the source plane. I will discuss all of this in far greater
detail In Chapter~\ref{ch:microlensing} but for now I should mention a few key
points:
\begin{itemize}
    \item Every point inside the two images maps to a single point on the source disc via
          the lens equation. The inverse mapping is one-to-many because each point on the
          source disc maps to two points (point lens images) in the image plane. The
          points on the limbs of the two images correspond to the points on the limb of
          the source disc. We cannot predict the location of the images without inverting
          the lens mapping
    \item To numerically compute the magnification of an extended source, we could either
          integrate the magnification function convolved with the source brightness
          profile over the entire source disc (a two-dimensional integral), or we could
          do the same integral in the image plane. The first approach is highly
          numerically unstable because we would need to integrate over a function which
          diverges as $1/r$ approaching the caustic and this would require a very large
          number of lens equation evaluations. Hence it is much better to integrate in
          the image plane where the function is smooth.
    \item Decreasing the source radius $\rho_\star$ results in narrower arcs of the
          images (smaller radial extent).
\end{itemize}

\subsection{Parallax}
In Equation~\ref{eq:relative_trajectory_no_parallax} we have assumed that the
relative motion of the lens with respect to the source on the plane of the sky
is rectilinear. This is a reasonable approximation in a barycentric frame of
reference if the acceleration of the source and the lens can be neglected.
However, most microlensing events are observed from a non-inertial geocentric
frame so that the observed trajectory is modified due the local acceleration of
the Earth and this effect needs to be taken into account in a realistic model.
In order to include the parallax effect, we need to modify
Equations~\ref{eq:source_position} and \ref{eq:lens_position} to include the
projected motion of the Sun relative to the Earth on the plane of the sky. We
project $\mathbf s$ -- the position vector of the Sun relative to Earth, onto a
plane perpendicular to the line of sight towards the source star (plane of the
sky) which is defined by the unit vector $\boldsymbol{\hat n}$ normal to the
plane. The unit vector $\boldsymbol{\hat n}$ depends on the sky coordinates of
the source star $(\alpha,\delta)$ (right ascension and declination).

We choose to work in geocentric equatorial coordinates defined by the spherical
unit vectors $\hat{\mathbf e}_n$, pointing North, and $\hat{\mathbf e}_e$
pointing eastward such that the coordinate system is right-handed. The unit
vectors are defined by
\begin{align}
    \hat{\mathbf e}_e & = \hat{\mathbf z}\times \hat{\mathbf n}  \\
    \hat{\mathbf e}_n & = \hat{\mathbf n}\times\hat{\mathbf e}_e
\end{align}
The two components of $\mathbf s$ projected onto the plane of the sky are then
\begin{align}
    \zeta_e(t;\alpha,\delta)\equiv \mathbf s\cdot \hat{\mathbf e}_e \\
    \zeta_n(t;\alpha,\delta)\equiv \mathbf s\cdot \hat{\mathbf e}_n
\end{align}
In practice, we can retrieve $\mathbf{s}$ using NASA's JPL Horizons system and compute
the projected separation at any time $t$.
The angular positions of the source and the lens
(Equations~\ref{eq:source_position} and \ref{eq:lens_position}) then become
\begin{align}
    \boldsymbol{\theta}_S(t) & = \boldsymbol{\theta}_{S,0}+(t-t_0)\boldsymbol{\mu}_S
    +\pi_S\,\boldsymbol{\zeta}(t)                                                    \\
    \boldsymbol{\theta}_L(t) & = \boldsymbol{\theta}_{L,0}+(t-t_0)\boldsymbol{\mu}_L
    +\pi_L\,\boldsymbol{\zeta}(t)
\end{align}
where  $\pi_S\equiv 1\,\mathrm{au}/D_S$ is the source parallax, and $\pi_L$ is the
lens parallax.
The relative separation is then
\begin{equation}
    \boldsymbol{u}(t)= \frac{\boldsymbol\theta_{LS, 0}}{\theta_E}
    +\frac{t-t_0}{\theta_E}\boldsymbol{\mu}_{LS}+\pi_{E}\,\boldsymbol{\zeta}(t)
    \label{eq:relative_separation_parallax}
\end{equation}
with $\pi_E\equiv\pi_{LS}/\theta_E\equiv (\pi_L-\pi_S)/\theta_E$.

Since annual parallax is a higher order effect affecting the apparent
trajectory of the lens on the sky, it makes sense to decompose the trajectory
as a sum of rectilinear motion plus a deviation due to parallax. Following
\citet{2004ApJ...606..319G} we define the position offset of the Sun on the
plane of the sky relative to its position at a particular time $t_0'$ as

\begin{equation}
    \delta\boldsymbol \zeta (t)=\boldsymbol \zeta (t)-\boldsymbol \zeta (t_0')-(t-t_0')
    \boldsymbol{\dot \zeta} (t_0')
\end{equation}
By construction, we have $\delta\boldsymbol \zeta (t_0')=0$ and
$\delta\dot{\boldsymbol \zeta} (t_0')=0$.
Equation~\ref{eq:relative_separation_parallax} then takes the form
\begin{equation}
    \boldsymbol{u}(t)=\mathbf{u}_0 + (t-t_0')\,\dot{\mathbf{u}}_0 +
    \pi_E\,\delta\boldsymbol \zeta(t)
    \label{eq:relative_separation_parallax_decomposed}
\end{equation}
where
\begin{align}
    \mathbf{u}_0       & \equiv \mathbf{u}(t_0) =\frac{\boldsymbol\theta_{LS, 0}}{\theta_E}
    + \frac{(t_0'-t_0)}{\theta_E}\boldsymbol\mu_{LS}+
    \pi_E\,\boldsymbol \zeta(t_0)                                                           \\
    \dot{\mathbf{u}}_0 & \equiv\dot{\mathbf{u}}(t_0)=
    \frac{\boldsymbol\mu_{LS}}{\theta_E}  + \pi_E\,\dot{\boldsymbol \zeta}
    (t_0')
\end{align}
In what follows, we set the reference time to $t_0'\equiv t_0$ in order to ensure that
the vectors $\mathbf{u}_0$ and $\dot{\mathbf{u}}_0$ are perpendicular to each other,
which is the case at $t=t_0$ because at $t_0$ the trajectory has a local extremum
\footnote{In CITE, $t_0'$ is fixed to a particular value. It is not clear
    to me how this ensures that the two vectors are perpendicular to each other
    (TODO: show that this is true).}

\subsubsection{$\mathbf{u}(t)$ in the $(\mathbf{\hat e}_\bot,
        \mathbf{\hat e}_\parallel)$ coordinate system}%

\begin{figure}
    \centering
    \includegraphics[width=0.3\linewidth]{static/microlensing/microlensing_parallax_coordinates.pdf}
    \caption{A coordinate system
        $(\mathbf{\hat e}_\bot,\mathbf{\hat e}_\parallel)$ parallel to the lens trajectory at time $t_0$. The orange curve
        represents the lens trajectory relative to the source star at the origin. The
        coordinate system $(\mathbf{\hat e}_\bot,\mathbf{\hat e}_\parallel)$ is related to the equatorial coordinates $(\mathbf{\hat
                e}_e,\mathbf{\hat e}_n)$ by a rotation through an angle $\psi$.}
    \label{fig:parallax2}
\end{figure}
